\addchap{\lsAcknowledgementTitle} 

\begin{refsection}

We would like to express our gratitude to all who have contributed to this Festschrift for Daniel Everett. 

First and foremost, we would like to thank the publisher, Language Science Press, for their professional support and guidance throughout this project. We acknowledge that Language Science Press do not publish festschrifts, and that a special exception was made here. We extend a special thank you to Stefan Müller and Sebastian Nordhoff, who have dedicated an unusual amount of attention and assistance to this festschrift. Unfortunately, this festschrift supports Müller's Festschrift Universal stating that festschrifts require an enormous amount of work (\citealt[xvi]{chapters/0_mueller}, this volume). I am glad that Language Science Press has managed to publish the book in a finite amount of time.

We also wish to thank all the speakers, in-person and virtual, who presented their research in the event on June 8th at MIT, \textit{From Fieldwork to Linguistic Theory: A Tribute to Dan Everett}, which was fully recorded and can now be found on Youtube. We also express special gratitude to those who, in addition to presenting their work in the event, contributed their manuscripts to this publication. This festschrift would not exist without you. 

Our thanks go to the McGovern Institute and MIT for hosting and filming the event. We wish to thank Elise Malvicini from the McGovern Institute for her organizational efforts. Her work ensured the event ran smoothly. 

Finally, the first editor would like to extend a special thank you to Ev Fedorenko, for help with the organization behind the scenes, and for emotional support always.

To all, your contributions have been invaluable in making this Festschrift a fitting tribute to Daniel Everett.

\printbibliography[heading=subbibliography,notkeyword=this]

\end{refsection}